### 2018-10-09 manuscript.Rnw
almost resubmitted

\section*{Design and Implementation}
\texttt{rethomics} is implemented in \texttt{R}\cite{r_core_team_r_2017}
as a collection of packages (Fig~\ref{fig:fig-1}).
Such modular architecture follows the model of modern frameworks such as the \texttt{tidyverse}\cite{wickham_tidyverse_2017}, which results in increased testability, maintainability and adaptability.
In this model, each task of the analysis workflow (\emph{i.e.} data import, manipulation and visualisation) is handled by a different package, and new ones can be designed to suit specific needs.
At the core of \texttt{rethomics} lies the \texttt{behavr} table, a structure used to store large amounts of data (\emph{e.g.} position and activity) and metadata (\emph{e.g.} treatment and genotype) in a unique \texttt{data.table}-derived object\cite{dowle_data.table_2017}.
Any input package will import experimental data as a \texttt{behavr} table which can, in turn, be analysed and visualised regardless of the original input platform.
Numerical results and plots are standard objects that can, therefore, be further analysed inside the wide \texttt{R} package ecosystem.

@@

% We present an small example examining the circadian behaviour of 128 fruit flies in recorded in a DAM2 -- a paradigm very widely adopted. 
% A less formal description of this case and others are explained at \href{https://rethomics.github.io/}{https://rethomics.github.io/}.
% In order to prodive a more comprehensive and didactic example, data was altered and simplified.	
% Fig~\ref{fig:experiment} describes our case experiment (A) and the corresponding metadata (B).
% Briefly, three monitors were used each containing 16 males and 16 females of a different genotype (A, B and C).
% Two replicates were performed at different times.
% raw data and metadata files are publicly available TODO cite zenodo
%

\section*{Availability and Future Directions}
All packages in the \texttt{rethomics} framework are available under the terms of the GPLv3 license and are listed at
\href{https://rethomics.github.io/intro.html}{https://rethomics.github.io/intro.html}.
\texttt{behavr} and \texttt{damr} are already hosted on the Comprehensive R Archive Network (CRAN), and all the other packages are either under review or in preparation.
Extensive installation instructions, as well as reproducible demos and tutorials, are available at
\href{https://rethomics.github.io/}{https://rethomics.github.io/}.
All packages are continuously integrated and unit tested on several versions of \texttt{R} to minimise the risk of present and future issues.

Several users, in different research groups, have already adopted and are contributing to the future development framework.
Several new packages in the \texttt{rethomics} framework are currently envisaged. 
They include utilities to input new behaviour tracking methods and analyse multi-animal interactions.

>>>>

\section*{Design and Implementation}
\texttt{rethomics} is implemented in \texttt{R}\cite{r_core_team_r_2017}
as a collection of packages (Fig~\ref{fig:fig-1}).
A summary of the functionalities of all packages is represented in a supplementary table (\nameref{S1-Table}).
Such modular architecture follows the model of modern frameworks such as the \texttt{tidyverse}\cite{wickham_tidyverse_2017}, which results in increased testability, maintainability and adaptability.
In this model, each task of the analysis workflow (\emph{i.e.} data import, manipulation and visualisation) is handled by a different package, and new ones can be designed to suit specific needs.
At the core of \texttt{rethomics} lies the \texttt{behavr} table, a structure used to store large amounts of data (\emph{e.g.} position and activity) and metadata (\emph{e.g.} treatment and genotype) in a unique \texttt{data.table}-derived object\cite{dowle_data.table_2017}.
Any input package will import experimental data as a \texttt{behavr} table which can, in turn, be analysed and visualised regardless of the original input platform.
Numerical results and plots are standard objects that can, therefore, be further analysed inside the wide \texttt{R} package ecosystem.

@@

% We present an small example examining the circadian behaviour of 128 fruit flies in recorded in a DAM2 -- a paradigm very widely adopted. 
% A less formal description of this case and others are explained at \href{https://rethomics.github.io/}{https://rethomics.github.io/}.
% In order to prodive a more comprehensive and didactic example, data was altered and simplified.	
% Fig~\ref{fig:experiment} describes our case experiment (A) and the corresponding metadata (B).
% Briefly, three monitors were used each containing 16 males and 16 females of a different genotype (A, B and C).
% Two replicates were performed at different times.
% raw data and metadata files are publicly available TODO cite zenodo

\section*{Discussion}

The framework we presented in this article describes a general computational set of tools to store, manipulate and analyse behavioural data. 
As a practical case study, in this article, we showed how \texttt{rethomics} can readily be used to analyse circadian and sleep data.
As such, it already fills a clear gap in the sleep and circadian community by providing an open, extensible and programmatic toolbox as opposed to a rigid array of GUI tools which cannot interoperate. 
In particular, within the R environment, rethomic features a streamlined workflow (Fig.~\ref{fig:fig-1}): data loading; quality control (Fig.~\ref{fig:fig-3}); per-individual visualisation (Fig.~\ref{fig:fig-4}A) and computation (Fig.~\ref{fig:fig-4}B); group averages and statistical analysis (Fig.~\ref{fig:fig-5}), ultimately resulting in publication-quality figures. 
In addition to the canonical analysis we presented in this study, the richness of the \texttt{R} environment makes \texttt{rethomics} particularly interesting for exploratory work: as a case in point, we showed here the use of continuous wavelet transform to study the modulation of walking pace (Fig.~\ref{fig:fig-6}). Altogether, \texttt{rethomics} will allow the community to share solutions for the analysis and visualisation of behavioural data originated from a plethora of recording equipment. 
By design, \texttt{rethomics} is meant to be flexible and certainly not limited to the study of circadian rhythms.
In fact, \texttt{behavr} tables (Fig.~\ref{fig:fig-2}) are a general structure to link data to metadata that come with no assumption regarding the type and number of variables stored.
Therefore, we anticipate it will apply to a variety of contexts and scales.
In the last decade, multiple devices have been developed -- either commercially or academically -- to record longitudinal data on several animal models (i.e. feeding, metabolic activity, heart rate, position and so on).
All these devices and behavioural paradigm could take advantage of the skeletal structure in \texttt{rethomics} and lead to  the development of device-specific packages.
Such an approach becomes particularly useful in order to compare methods or study behaviours that are recorded by different devices on the same individual animals.

\section*{Availability}
All \texttt{rethomics} packages are listed at \href{https://rethomics.github.io/intro.html}{https://rethomics.github.io/intro.html}.
and are hosted on the official Comprehensive \texttt{R} Archive Network (CRAN) under the terms of the GPLv3 license. 
Installation instructions, as well as reproducible demos and tutorials, are available at
\href{https://rethomics.github.io/}{https://rethomics.github.io/}.
All packages are continuously integrated and unit tested on several versions of \texttt{R} to minimise the risk of present and future issues.
